---
title: "scChIX example: performing model selection, unmixing double-incubated cells, and linking UMAPs"
output:
  md_document:
    variant: markdown_github
vignette: >
  %\VignetteIndexEntry{scChIX-simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  cache  = TRUE,
  fig.width = 14,
  fig.width = 14,
  comment = "#>"
)
```




```{r setup, echo=TRUE, include=TRUE}
library(scChIX)
library(dplyr)
library(ggplot2)
library(data.table)
library(umap)
library(hash)
library(igraph)
# library(devtools)
# install_github("bowang-lab/simATAC")
library(simATAC)
library(JFuncs) # for plotting multiple panels in one plot
library(irlba)
```


## Simulating scChIX input data 


First we set up some parameters to simulate scChIX data. We will set half the bins to be "mutually exclusive" while the other half will be overlapping.

```{r init, out.width="80%", echo=TRUE}

hubprefix <- getwd()
outdir <- file.path(hubprefix, "snakemake_inputs_countmats")
dir.create(outdir, recursive = TRUE)

jspec <- "hg38"
jseed <- 0
jncells.per.rep <- 250
jnbins <- 10000
jlibmean <- 12
jlibsd <- 1
jlibp <- 0.5
jzp <- 1
shuffle.rows.seed <- jseed

sim.params <- list(jspec = jspec,
                   jseed = jseed,
                   jncells.per.rep = jncells.per.rep,
                   jnbins = jnbins,
                   jlibmean = jlibmean,
                   jlibsd = jlibsd,
                   jlibp = jlibp,
                   jzp = jzp)


# Run simulation ----------------------------------------------------------

ctypes <- c("A", "B", "C")
names(ctypes) <- ctypes

ctype.params <- list(ctype.name = c("A", "B", "C"),
                     jseed = c(0, 123, 999),
                     shuffle.rows.seed = c(0, 123, 999),
                     frac.mutual.excl = c(0.5, 0.5, 0.5))

ctype.params.lst <- lapply(ctypes, function(ctype){
  i <- which(ctype.params$ctype.name == ctype)
  return(list(ctype = ctype,
              jseed = ctype.params$jseed[[i]],
              shuffle.rows.seed = ctype.params$shuffle.rows.seed[[i]],
              frac.mutual.excl = ctype.params$frac.mutual.excl[[i]]))
})
```

For each of the three cell types, we generate three count matrices. First represents histone mark 1, second represents histone mark 2, and the third represents histone mark 1+2 double-incubation.

```{r generate_counts, out.width="80%", echo=TRUE}


ctype.sim.counts.lst <- lapply(ctypes, function(jctype){
  scChIX::SimulateChICseq(ctype.name = jctype,
                  jseed = ctype.params.lst[[jctype]]$jseed,
                  shuffle.rows.seed = ctype.params.lst[[jctype]]$shuffle.rows.seed,
                  frac.mutual.excl = ctype.params.lst[[jctype]]$frac.mutual.excl)
})


# Create simulated matrix -------------------------------------------------


jmark1 <- "mark1"
jmark2 <- "mark2"
jmarkdbl <- paste(jmark1, jmark2, sep = "-")

mat.mark1 <- scChIX::ConcatMat(ctype.sim.counts.lst, ctypes, jmark = "mark1")
mat.mark2 <- scChIX::ConcatMat(ctype.sim.counts.lst, ctypes, jmark = "mark2")
mat.markdbl <- scChIX::ConcatMat(ctype.sim.counts.lst, ctypes, jmark = "markdbl")

mat.mark.lst <- list(mat.mark1, mat.mark2, mat.markdbl)
names(mat.mark.lst) <- c(jmark1, jmark2, jmarkdbl)

jmarks <- names(mat.mark.lst)
names(jmarks) <- jmarks

```

We do a quick check the UMAPs of these count matrices that they are three different cell types for each histone mark. 

```{r check, out.width="80%", echo=TRUE}


jsettings <- umap.defaults
jsettings$n_neighbors <- 150
jsettings$min_dist <- 0.1
jsettings$random_state <- 123
cbPalette <- c("#696969", "#56B4E9", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#006400",  "#32CD32", "#FFB6C1", "#0b1b7f", "#ff9f7d", "#eb9d01", "#2c2349", "#753187", "#f80597")

dat.umap.init.lst <- lapply(mat.mark.lst, function(jmat){
  print(head(jmat[1:5, 1:5]))
  lsi.out <- scchicFuncs::RunLSI(as.matrix(jmat)) # requires irlba
  dat.umap.mark <- scchicFuncs::DoUmapAndLouvain(topics.mat = lsi.out$u, jsettings = jsettings)
  return(dat.umap.mark)
})

jmarks <- names(dat.umap.init.lst); names(jmarks) <- jmarks
m.lst <- lapply(jmarks, function(jmark){
  jdat <- dat.umap.init.lst[[jmark]]
  m <- ggplot(jdat, aes(x = umap1, y = umap2, color = louvain)) +
    geom_point() +
    theme_bw() +
    ggtitle(jmark) + 
    scale_color_manual(values = cbPalette) +
    theme(aspect.ratio=1, panel.grid.major = element_blank(), panel.grid.minor = element_blank())
  return(m)
})

print(m.lst)

```

## Running scChIX on input data

Write the count matrices to a directory (see `inst/extdata/countmat_var_filt.mark1.rds` 
for example) and then run the snakemake workflow 
`snakemake_workflow/run_snakemake.simulation_data.sh` changing the paths to 
the correct directories. 

The snakemake workflow takes several hours to complete, so we will just load
and analyze the results in this notebook. 


## Downstream analysis of scChIX to check simulated data 

Load the scChIX outputs for the simulated data for three overlapping scenarios:
`frac.mutual.excl=0.01, 0.5, 0.99`

```{r load_outputs, out.width="80%", echo=TRUE}
data(Simulation_scChIX_Outputs, verbose=FALSE)
```

We can plot the empirical 95% confidence intervals from our estimates of
the degree of overlaps.


```{r ci, out.width="80%", echo=TRUE}


# we use the scenario of 0.99 mutually exclusive bins because that spans 
# the range of degree of overlaps from 0 to 1.
dat.binmeans.binned <- dat.binmeans.lst.lst$`0.99` %>%
  bind_rows() %>%
  rowwise() %>%
  mutate(xbin = round(BinMean.dbl, 2)) %>%
  group_by(xbin) %>%
  summarise(Mean = mean(overlap.estimate),
            StdDev = sd(overlap.estimate),
            CI.lower = quantile(overlap.estimate, 0.05),
            CI.upper = quantile(overlap.estimate, 0.95),
            CI.lower.centered = CI.lower - Mean,
            CI.upper.centered = CI.upper - Mean)

ggplot(dat.binmeans.binned, aes(x = xbin, y = StdDev)) +
  geom_point() +
  theme_bw() +
  theme(aspect.ratio=1, panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  xlab("Fraction of Bin Signal Multiplexed to Mark 1") +
  ylab("Standard Deviation")

ggplot(dat.binmeans.binned, aes(x = xbin, y = Mean, ymin = CI.lower, ymax = CI.upper)) +
  geom_abline(slope = 1, color = 'blue', size = 1, alpha = 0.5) +
  geom_point() +
  geom_errorbar() +
  theme_bw() +
  theme(aspect.ratio=1, panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  xlab("True Fraction of Bin Signal Multiplexed to Mark 1") +
  ylab("Inferred Fraction of Bin Signal Multiplexed to Mark 1\n(ErrorBars: 95% CI)")

ggplot(dat.binmeans.binned, aes(x = xbin, y = 0, ymin = CI.lower.centered, ymax = CI.upper.centered)) +
  geom_point() +
  geom_errorbar() +
  theme_bw() +
  theme(aspect.ratio=1, panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  xlab("True Fraction of Bin Signal Multiplexed to Mark 1") +
  ylab("95% Confidence Interval (+/- Fraction)") +
  coord_cartesian(ylim = c(-0.1, 0.1))

ggplot(dat.binmeans.binned, aes(x = xbin, y = 0, ymin = CI.lower.centered, ymax = CI.upper.centered)) +
  geom_point() +
  geom_errorbar() +
  theme_bw() +
  theme(aspect.ratio=0.33, panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  xlab("True Fraction of Bin Signal Multiplexed to Mark 1") +
  ylab("95% Confidence Interval (+/- Fraction)") +
  coord_cartesian(ylim = c(-0.1, 0.1))


```

We can show that the two UMAPs can be linked together using the deconvolved
double-incubated cells as anchors.


```{r umap, out.width="80%", echo=TRUE}

# Plot umaps  -------------------------------------------------------------

dat.umap.long <- dat.umap.lst %>% bind_rows() %>%
  rowwise() %>%
  mutate(umap2.scale = ifelse(ctype == "B", umap2.scale + 0.75, umap2.scale)) # shift up to improve visualization

cbPalette.ctype <- c("#FFB6C1", "#32CD32", "#56B4E9", "#FFB6C1", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#006400", "#FFB6C1", "#32CD32", "#0b1b7f", "#ff9f7d", "#eb9d01", "#7fbedf")
ggplot(dat.umap.long, aes(x = umap1.scale.shift, y = umap2.scale, color = ctype, group = cell)) +
  geom_point() +
  geom_path(alpha = 0.02) +
  geom_vline(xintercept = 0, linetype = "dotted") +
  scale_color_manual(values = cbPalette.ctype, na.value = "grey85") +
  facet_wrap(~frac.overlapping.bins) +
  theme_bw() +
  theme(aspect.ratio=0.5, panel.grid.major = element_blank(), panel.grid.minor = element_blank(), legend.position = "bottom")

```


Finally we can plot the distribution of overlap estimates across the bins and
compare how these distributions look compared to ground truth.


```{r hist, out.width="80%", echo=TRUE}
frac.mutexcl.str.vec <- c("0.01", "0.5", "0.99"); names(frac.mutexcl.str.vec) <- frac.mutexcl.str.vec

dat.binmeans.lst <- lapply(frac.mutexcl.str.vec, function(jfrac){
  dat.binmeans.lst.lst[[jfrac]] %>%
    bind_rows()
})

m.lst <- lapply(frac.mutexcl.str.vec, function(jfrac){
  jtitle <- jfrac # mutually exclusive
  jtitle <- 1 - as.numeric(jtitle) # degree of overlap
  m <- ggplot(dat.binmeans.lst[[jfrac]], aes(x = BinMean.dbl, fill = annot, y = ..count../sum(..count..))) +
    geom_histogram(bins = 50, position = "identity", alpha = 0.5) +
    ggtitle(paste("True Fraction of Signal Multiplexed to Mark 1. Frac overlap: ", jtitle)) +
    theme_bw() +
    theme(aspect.ratio=1, panel.grid.major = element_blank(), panel.grid.minor = element_blank(), legend.position = "bottom")
  return(m)
})

print(m.lst)
JFuncs::multiplot(m.lst[[1]], m.lst[[2]], m.lst[[3]], cols = 3)


m.infer.lst <- lapply(frac.mutexcl.str.vec, function(jfrac){
  jtitle <- ifelse(jfrac == "1.0", "0.99", jfrac)
  jtitle <- 1 - as.numeric(jtitle)
  m <- ggplot(dat.binmeans.lst[[jfrac]], aes(x = overlap.estimate, fill = annot, y = ..count../sum(..count..))) +
    geom_histogram(bins = 50, position = "identity", alpha = 0.5) +
    ggtitle(paste("Inferred Fraction of Signal Multiplexed to Mark 1. Frac overlap: ", jtitle)) +
    theme_bw() +
    theme(aspect.ratio=1, panel.grid.major = element_blank(), panel.grid.minor = element_blank(), legend.position = "bottom")
  return(m)
})
print(m.infer.lst)
JFuncs::multiplot(m.infer.lst[[1]], m.infer.lst[[2]], m.infer.lst[[3]], cols = 3)


```
